---
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook: default
  html_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    highlight: tango
fontsize: 12pt
geometry: margin = 1.2in
header-includes:
- \usepackage{placeins}
- \usepackage{fancyhdr}
- \usepackage{setspace}
- \usepackage{chngcntr}
- \usepackage{microtype}
always_allow_html: yes
urlcolor: blue
---

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, comment = "#>")
library(png)
library(grid)
library(gridExtra)
```

# Installing cerUB {#install}

There are two options for installing the *cerUB* package:

a. Downloading the latest release version from [Zenodo.org](zenodo.org) and installing it in [RStudio](rstudio.com) (Tools > Install Packages... > Install from: Package Archive File).

> Angourakis, Andreas, & Marti&#769;nez Ferreras, Vero&#768;nica. (2017, September 23). cerUB - Protocols for exploring archaeometric data (R package). Zenodo. http://doi.org/10.5281/zenodo.975451

b. Installing the latest development version directly from [GitHub](github.com) ([Andros-Spica/cerUB](https://github.com/Andros-Spica/cerUB)), using the *devtools* package:

```{r, eval=FALSE}
# this will install devtools package, if not installed already
if (!require("devtools"))
    install.packages("devtools") 

devtools::install_github("Andros-Spica/cerUB")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# this will install devtools package, if not installed already
if (!require("devtools"))
    install.packages("devtools") 

if (!require("cerUB"))
    devtools::install_github("Andros-Spica/cerUB")
if (!require("biplot2d3d"))
    devtools::install_github("Andros-Spica/biplot2d3d")
```

The second option is recommended, because it is a faster way to install and update packages that are not in [CRAN](cran.r-project.org).


The same options are available for installing the *biplot2d3d* package, which we use here to plot protocols results.

> Andreas Angourakis. (2017, September 20). biplot2d3d - an R package for generating highly-customizable biplots. Zenodo. http://doi.org/10.5281/zenodo.897603

```{r, eval=FALSE}
devtools::install_github("Andros-Spica/biplot2d3d")
```

Any other package required by these two packages (*ade4*, *rgl*, etc.) should be automatically installed.
